---
title: "Assignment 4"
author: "Emil Hafeez (eh2928)"
date: "10/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library("animation")

theme_set(theme_minimal() + theme(legend.position = "bottom")) #setup and establish the colors schemes for plits
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
```{r}
knee_df =
  read_csv(
      "./data/Knee.csv") #load in df
```

# Problem 1

In the context of ANOVA models, prove the partitioning of the total variability (sum of squares).

\begin{equation}
\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(y_{i j}-\overline{\bar{y}}\right)^{2}=\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(y_{i j}-\bar{y}_{l}\right)^{2}+\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(\bar{y}_{i}-\overline{\bar{y}}\right)^{2}
\end{equation}

We assume that the samples are drawn independently from the underlying populations, that the distributions of the error terms are normal, and that the variances of the k populations are equal (and thus that the variance of the outcome does not depend on the sample). 

Let \(n\) be the total number of observations

Now, we can write that the total variation is equal to \(\sum_{i=1}^{n}(y_i - \overline{y})^2\), giving the sum of squares for these data where \(y_i\) is the ith data point, and where \(\bar{y}\) is the estimate of the mean. Now, more specifically, 

Let \(\bar{y}\) be the mean from each of the  the $i^{th}$ group where $i$ = 1,2,3,...\(k\) is the number of groups and let \(\overline{\bar{y}}\) be the grand mean, such that \(\overline{\bar{y}}=\frac{\sum_{i=1}^{k} \sum_{j=1}^{n_{i}} y_{i j}}{n}\). 




