---
title: "Assignment 4"
author: "Emil Hafeez (eh2928)"
date: "10/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library("animation")

theme_set(theme_minimal() + theme(legend.position = "bottom")) #setup and establish the colors schemes for plits
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
```{r}
knee_df =
  read_csv(
      "./data/Knee.csv") #load in df
```

# Problem 1

In the context of ANOVA models, prove the partitioning of the total variability (sum of squares).

\begin{equation}
\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(y_{i j}-\overline{\bar{y}}\right)^{2}=\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(y_{i j}-\bar{y}_{l}\right)^{2}+\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(\bar{y}_{i}-\overline{\bar{y}}\right)^{2}
\end{equation}

We assume that the samples are drawn independently from the underlying populations, that the distributions of the error terms are normal, and that the variances of the k populations are equal (and thus that the variance of the outcome does not depend on the sample). 

Let \(n\) be the total number of observations

Let \(\bar{y}\) be the mean from the $i^{th}$ group

Let \(\bar{y_i}\) be the grand mean, such that \(\overline{\bar{y}}=\frac{\sum_{i=1}^{k} \sum_{j=1}^{n_{i}} y_{i j}}{n}\)

Let \(\overline{\bar{y}}\)
