---
title: "Assignment 4"
author: "Emil Hafeez (eh2928)"
date: "10/28/2020"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library("animation")
library(multcomp)

theme_set(theme_minimal() + theme(legend.position = "bottom")) #setup and establish the colors schemes for plits
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```
```{r, include = F}
knee_df =
  read_csv(
      "./data/Knee.csv") #load in df
```

# Problem 1 (5p)

In the context of ANOVA models, prove the partitioning of the total variability (sum of squares).

\begin{equation}
\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(y_{i j}-\overline{\bar{y}}\right)^{2}=\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(y_{i j}-\bar{y}_{l}\right)^{2}+\sum_{i=1}^{k} \sum_{j=1}^{n_{i}}\left(\bar{y}_{i}-\overline{\bar{y}}\right)^{2}
\end{equation}

We assume that the samples are drawn independently from the underlying populations, that the distributions of the error terms are normal, and that the variances of the k populations are equal (and thus that the variance of the outcome does not depend on the sample). 

Let \(n\) be the total number of observations

Now, we can write that the total variation is equal to \(\sum_{i=1}^{n}(y_i - \overline{y})^2\), giving the sum of squares for these data where \(y_i\) is the ith data point, and where \(\bar{y}\) is the estimate of the mean. Now, more specifically, 

Let \(\bar{y}\) be the mean from each of the  the $i^{th}$ group where $i$ = 1,2,3,...\(k\) is the number of groups and let \(\overline{\bar{y}}\) be the grand mean, such that \(\overline{\bar{y}}=\frac{\sum_{i=1}^{k} \sum_{j=1}^{n_{i}} y_{i j}}{n}\). 

....Continue




# Problem 2 

## Part a)
Generate descriptive statistics for each group and comment on the differences observed. (4p)

Descriptives: let's do mean for each group and sd, range, and median for each group, and comment.

```{r, include = F}
knee_df %>% 
  pivot_longer(
  Below:Above,
  names_to = "group",
      values_to = "days_to_recover"
  ) %>% 
    group_by(group) %>% 
  mutate(
    mean = mean(days_to_recover, na.rm = TRUE),
    median = median(days_to_recover, na.rm = TRUE),
    sd = sd(days_to_recover, na.rm = TRUE)
  )

knee_tidied_df =
knee_df %>% 
  pivot_longer(
  Below:Above,
  names_to = "group",
      values_to = "days_to_recover"
  ) %>% 
    group_by(group) %>% 
  mutate(
    mean = mean(days_to_recover, na.rm = TRUE),
    median = median(days_to_recover, na.rm = TRUE),
    sd = sd(days_to_recover, na.rm = TRUE)
  )
```
Regarding participants' physical status before therapy, the "Above Average" group had a mean of 23.57 days (standard deviation = 4.20), and a median of 22 days.  The "Average" group had a mean of 33 days to recover (standard deviation = 3.92 days), and a median of 32 days. The "Below Average" group had a mean of 38 days to recover (standard deviation = 5.478 days) and a median of 40 days to recover. Thus, it appears that worse prior physical status is associated with more days to recovery. Interestingly, it also appears that a worse prior physical status seems to have higher dispersion (as per the SD, as well). Finally, it appears visually that only the "Below" group had left skew, whereas the other two groups had right skew. This can be explored visually, as below.

```{r, echo = F, warning = F}
knee_tidied_df %>% 
  arrange(group, desc(group)) %>% 
  ggplot(aes(x = group, y = days_to_recover)) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom="point", shape=20, size=5, color="red", fill="red") +
  labs(
    title = "Rehabilitation Days to Recovery By Prior Physical Status",
    x = "Days to Recover",
    y = "Prior Physical Status Group",
    caption = "Figure 1, Biostatistics P8130 Assignment 4")
```

## Part b)

Let \(\alpha\) = 0.01

\(H_0: \mu_1 = \mu_2 = \mu_3 \), ie that all group means (the Below, Average, and Above groups respectively) are equal

\(H_A: \) at least two of the population means differ, ie not all group means are equal


\begin{tabular}{|l|l|l|l|l|}
\hline
Source & Sum of Square (SS) & Degrees of Freedom (df) & Mean Sum of Square & F-Statistic \\ \hline
Between & Between SS & k-1 & Between SS / (k-1) & F = \(\frac{(Mean Between SS / k-1)}{(Within SS / (n-k)}\) \\ \hline
Within & Within SS & n-k & Within SS / (n-k) &  \\ \hline
Total & Between SS + Within SS & n-1 &  &  \\ \hline
\end{tabular}

```{r, include = F}
res <- lm(days_to_recover~factor(group), data = knee_tidied_df)
res
anova(res)

res1<-aov(days_to_recover~factor(group), data=knee_tidied_df)
summary(res1)
```

\(F = \frac{(Mean Between SS / k-1)}{(Within SS / (n-k)}\) = 19.28

\(F_{critical} = \) `qf(0.95, 2, 22)` = 3.443357

Seeing as the decision rule is that we reject the null hypothesis if \(F > F_{k-1, n-k, 1-\alpha}\) and fail to reject the null hypothesis if \(F \leq F_{k-1, n-k, 1-\alpha}\), we find evidence to reject the null hypothesis and conclude that there is at least one group mean that is not equal to the other group means. In context, we find evidence to conclude that there is at least one difference in group mean recovery days to recovery until successful rehabilitation among the Below, Average, and Above Average prior physical fitness status. 


## Part c)

Seeing as we find evidence to reject the null hypothesis in the omnibus test, we can proceed to implementing a pairwise comparison method. We opt to perform each of three post-hoc comparisons, including the Bonferroni comparison, Tukey, and Dunnett.

Dear Dr. Lebron James,

I am writing to you to summarize the results of my analysis regarding your rehabilitation study. In order to investigate whether there are differences between the mean rehabilitation days to recovery time between the specified Below Average prior physical status, the Average prior physical status, and Above Average physical status groups, we implemented an ANOVA model and found evidence to reject the null hypothesis and conclude there is at least one difference between the group means \(F = \frac{(Mean Between SS / k-1)}{(Within SS / (n-k)} = 19.28\) and \(F_{critical} = 3.443357 \). Then, we implemented pairwise comparisons using several different methods, and found converging evidence in order to support a significant difference between the mean recovery time between the Average and Above Average group means, and between the Below Average and Above Average group means.

# Problem 2 

## Part a)

We consider a situation in which we are examining proportions, not means. McNemar's test for binomial proportions is not appropriate given we are not utilizing paired data, and one- or two-sample tests for binomial proportions are also not appropriate considering the number of categories we wish to compare. A Fisher's Exact Test is not feasible because this is not a 2x2 table.

Finally, considering that we are interested in testing whether distribution of swelling status is the same across the levels of the intervention group (vaccine / placebo), we opt to use a Chi-Squared test of homogeneity. 

## Problem b)


```{r, include = F}
vaccine_df <- matrix(c(54,42,134,16,32,142), nrow = 2, ncol = 3, byrow = T,
                  dimnames = list(c("Vaccine", "Placebo"), 
                                c("Major Swelling", "Minor Swelling", "No Swelling")))
vaccine_df                  
chisq.test(vaccine_df)
```

The table required to calculate the Chi Squared value is as follows.



\begin{tabular}{|l|l|l|l|l|}
\hline & \text { Major [Expected] } & \text { Minor [Expected] } & \text { No Swelling [Expected] } & \text { Total } \\
\hline \text { Vaccine } & 54 [33.33] & 42 [40.52] & 134 [171.14] & 230 \\
\hline \text { Placebo } & 16 [31.66] & 32 [33.48] & 142 [124.85] & 190 \\
\hline \text { Total } & 70 & 74 & 276 & 142 \\
\hline
\end{tabular}

using: 
```{r, echo = T, warning = F}
chisq.test(vaccine_df)$expected 
```


## Problem c)

We use \(\alpha = 0.05\) 

\(H_0 : p_{1,1} = p_{2,1} = p_{3,1}\), that the proportions of swelling category are equal among the vaccine status, AND that \(H_0 : p_{1,2} = p_{2,2} = p_{3,2}\) the proportion of swelling category are equal among the placebo status. 

The \(H_0\) is that not all proportions are equal.

The decision rule is that, at the 0.05 level, \(\chi^2 > \chi^2_{df,\alpha}\) where the degrees of freedom are given by \(df = (r-1)(3-1) = 2\) and the alpha has been set at 0.05. We fail to reject the null hypothesis if \(\chi^2 \leq \chi^2_{df,\alpha}\) . 

The test statistic is given by \(\chi^{2}=\sum_{i=1}^{2} \sum_{j=1}^{3} \frac{\left(O_{i j}-E_{i j}\right)^{2}}{E_{i j}}\) = \(26.082\). 

The critical value, as mentioned, is given by \(\chi^2_{(2-1)*(4-1), 0.95)}\) computed by `qchisq(0.95,2)` = 5.991465. 

Thus, we find evidence to reject the null hypothesis and conclude that there is sufficient evidence that the extent of swelling is significantly different by vaccine status.